Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/background.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

The ADT interface used here is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
presented in this report use \cc{unsigned integer}s as Items, which were
compared using a simple \n{less-than} function\footnote{This abstraction broke
down in several cases, such as where a \n{greater-than} function was required,
i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
word \n{key} is used to refer to items to be sorted.

@@

Sorts which are not comparison based also exist. These sorts are called counting
based sorts, as they count the number of keys to be put in each segment of an
array before putting them in place. The most important of these is radixsort,
discussed in Chapter \ref{radix}.

>>>>

The ADT interface used here is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
presented in this report use \n{unsigned integer}s as Items, which were
compared using a simple \n{less-than} function\footnote{This abstraction broke
down in several cases, such as where a \n{greater-than} function was required,
i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
word \n{key} is used to refer to items to be sorted.

@@

Sorts which are not comparison-based also exist. These sorts are called
counting-based sorts, as they count the number of keys to be put in each segment
of an array before putting them in place. The most important of these is
radixsort, discussed in Chapter \ref{radix}.

### 2005-08-14 docs/tech_report/conclusions.tex
fixed all references
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

LaMarca determines that radixsort could not be improved in the same was as the
comparison based sorts, and that due to poor cache performance, it ran more
slowly than quicksort. We showed, however, that one of the improvements applied
to mergesort could be applied to it, and that its resulting cache performance is
sufficiently good that its low instruction count and lack of branch
mispredictions make it the fastest sort in our survey.

@@

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\setcounter{figure}{8}
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption[Cycles per key of several major sorts and their variations]{\label{All cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

>>>>

LaMarca determines that radixsort could not be improved in the same was as the
comparison-based sorts, and that due to poor cache performance, it ran more
slowly than quicksort. We showed, however, that one of the improvements applied
to mergesort could be applied to it, and that its resulting cache performance is
sufficiently good that its low instruction count and lack of branch
mispredictions make it the fastest sort in our survey.

@@

\afterpage{
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-10em}
\setcounter{figure}{9} % note the cheating to make this work
\begin{figure}[H]
\begin{changemargin}
\includegraphics[angle=90,scale=1.8]{plots/all_cycles}
\vspace*{8em}
\end{changemargin}
\caption[Cycles per key of several major sorts and their variations]{\label{All
cycles}Cycles per key of several major sorts and their variations - this was
measured on a Pentium 4 using hardware performance counters.}
\end{figure}
\newpage
}

### 2005-08-14 docs/tech_report/heap.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

\section{Results}
All simulations presented here were performed on 32 byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 ints, we stick with 32 byte cache lines. \verb!UINT_MAX! was used
as the sentinel.

@@

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\cc{Creation, has children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\cc{Creation, child comparison}' branch is the comparison between the
two children in the 2-heap, and the `\cc{Creation, child promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\cc{Destruction, has
children}', `\cc{Destruction, child comparison}' and `\cc{Desruction child
promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\cc{fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

@@

The most interesting thing is the relation between the predictability of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsorts-3} and \ref{base and memory-tuned heapsorts-4} (the
8-heap).

>>>>

\section{Results}
All simulations presented here were performed on 32-byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \cc{int}s fit the cache
line.  LaMarca used 64 bit ints, and while this doesn't affect results, it means
that fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 ints, we stick with 32 byte cache lines. \cc{UINT\_MAX} was used
as the sentinel.

@@

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, has children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\n{Creation, child comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, child promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, has
children}', `\n{Destruction, child comparison}' and `\n{Desruction child
promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\n{fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

@@

The most interesting thing is the relation between the predictability of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\n{Destruction, child comparison}' and `\n{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsorts-1} (the 2-heap), with
the columns marked `\n{comparison 1}' to `\n{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsorts-3} and \ref{base and memory-tuned heapsorts-4} (the
8-heap).

### 2005-08-14 docs/tech_report/method.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

Once the algorithm was written, a testing framework was used. This tested for
edge-cases in the algorithms by testing sorts sized from $8$ to $255$ keys, with
different random data sets. Testing was also performed on a very large data set.
This test was timed and the results were considered. The results of this test
were not accurate: the program took up to a minute to run, during which time the
computer was being used by multiple users and processes, leading to context
switches, sharing of resources and scheduling. Despite these inaccuracies, it was
possible to use the results relative to each other, and to determine whether,
for example, memory-tuned quicksort was running faster than base quicksort.

@@

\subsection{SimpleScalar} \cc{SimpleScalar} is an architecture simulator. It
works by compiling code into its native instruction set, PISA, then emulating a
processor as it runs the program.  SimpleScalar can be found at
\n{http://www.simplescalar.com}. Documentation on how to use and modify
SimpleScalar are \cite{Burger97}, \cite{Austin02}, \cite{Tutorialv4},
\cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.

@@

\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of
data in two and three dimensions, and is used to create all of the graphs in
this document. Its homepage is \n{http://www.gnuplot.info/}.

@@

The main program provided the framework for the development of the sorts.
Imaginatively called \cc{main.c}, this program provided the means for testing
and (roughly) timing each sort. Figure \vref{Key sections of Main.c} contains a
sample of this code.

\code{Key sections of Main.c}{main.c}

\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

@@

Both our SimpleScalar simulations and our software predictor tests used truly
random data from \n{http://www.random.org}. The software predictor used ten
sections containing 4194304 \cc{unsigned integer}s, and the SimpleScalar
simulations used the same ten sections. When SimpleScalar simulations were run
on sized smaller than 4194304 keys, only the left hand side of the sections were
used.

\subsection{32-bit Integers}
LaMarca used 64-bit \cc{integer}s, as his test machine was a 64-bit system.
Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
architecture, LaMarca's results will not be identical to ours. Rather, the
number of cache misses in his results should be halved to be compared to ours;
when this is not the case, it is indicated.

>>>>

Once the algorithm was written, a testing framework was used. This tested for
edge-cases in the algorithms by testing sorts sized from \cc{8} to \cc{255}
keys, with different random data sets. Testing was also performed on a very
large data set.  This test was timed and the results were considered. The
results of this test were not accurate: the program took up to a minute to run,
during which time the computer was being used by multiple users and processes,
leading to context switches, sharing of resources and scheduling. Despite these
inaccuracies, it was possible to use the results relative to each other, and to
determine whether, for example, memory-tuned quicksort was running faster than
base quicksort.

@@

\subsection{SimpleScalar} \n{SimpleScalar} is an architecture simulator. It
works by compiling code into its native instruction set, PISA, then emulating a
processor as it runs the program.  SimpleScalar can be found at
\n{http://www.simplescalar.com}. Documentation on how to use and modify
SimpleScalar are \cite{Burger97}, \cite{Austin02}, \cite{Tutorialv4},
\cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.

@@

\subsection{Gnuplot}
\label{Gnuplot}
\n{Gnuplot} is a tool for plotting data files. It provides visualisations of
data in two and three dimensions, and is used to create all of the graphs in
this document. Its homepage is \n{http://www.gnuplot.info/}.

@@

The main program provided the framework for the development of the sorts.
Imaginatively called \cc{main.c}, this program provided the means for testing
and (roughly) timing each sort. Figure \vref{Key sections of main.c} contains a
sample of this code.

\code{Key sections of main.c}{main.c}

\subsection{Valgrind}
\label{Valgrind}
\n{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

@@

Both our SimpleScalar simulations and our software predictor tests used truly
random data from \n{http://www.random.org}. The software predictor used ten
sections containing 4194304 \n{unsigned integer}s, and the SimpleScalar
simulations used the same ten sections. When SimpleScalar simulations were run
on sized smaller than 4194304 keys, only the left hand side of the sections were
used.

\subsection{32-bit Integers}
LaMarca used 64-bit integers, as his test machine was a 64-bit system.
Since our machine is 32-bit, and SimpleScalar uses 32-bit addressing on our
architecture, LaMarca's results will not be identical to ours. Rather, the
number of cache misses in his results should be halved to be compared to ours;
when this is not the case, it is indicated.

### 2005-08-14 docs/tech_report/quick.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

Quicksort is an in-place, unstable sort. It was first described in
\cite{HoarePartition61} and \cite{HoareQuicksort61}, and later in more detail in
\cite{Hoare62}. Sedgewick did his PhD on quicksort, and several important
improvements to quicksort are due to this. They are discussed in
\cite{Sedgewick78} and \cite{Sedgewick02}. \cite{BentleyMcIlroy93} discusses
considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
equal keys and proves that quicksort has the lowest complexity of any comparison
based sort.

@@

Sedgewick's second improvement is the use of an elementary sort for small
sub-files.  Once a partition is smaller than an arbitrary threshold, insertion
sort is used to sort the partition. Sedgewick states that roughly the same
results are obtained for thresholds between 5 and 25. An additional reduction in
instruction count can be had by doing the insertion sort over the entire array
at the end, rather than across each small array as soon as the threshold is
reached. At the end of the partition sorting, LaMarca places a sentinel at the
end of the array and uses an insertion sort, eliminating the bounds check from
the end. In this implementation, the sentinel is placed at the front, and the
smallest of the first \cc{threshold} keys is used as a sentinel, in keeping
with our policy of not exceeding the bounds of the array.

@@

The number of keys in the linked list is 1022 since a word is needed for the
count, and another as a pointer to the next list. 1024 is the number of 32-bit
words which fit in a 4KB page of memory. In earlier versions where each page was
\cc{malloc}ed individually, the number of keys in the linked list was 1018,
as 4 words were used by the particular version of \cc{malloc} used by
SimpleScalar\footnote{This particular \cc{malloc} forms part of
\cc{glibc}, the GNU C library, version 1.09.}.

>>>>

Quicksort is an in-place, unstable sort. It was first described in
\cite{HoarePartition61} and \cite{HoareQuicksort61}, and later in more detail in
\cite{Hoare62}. Sedgewick did his PhD on quicksort, and several important
improvements to quicksort are due to this. They are discussed in
\cite{Sedgewick78} and \cite{Sedgewick02}. \cite{BentleyMcIlroy93} discusses
considerations for a fast C implementation. \cite{SedgewickBentley02} discusses
equal keys and proves that quicksort has the lowest complexity of any
comparison-based sort.

@@

Sedgewick's second improvement is the use of an elementary sort for small
sub-files.  Once a partition is smaller than an arbitrary threshold, insertion
sort is used to sort the partition. Sedgewick states that roughly the same
results are obtained for thresholds between 5 and 25. An additional reduction in
instruction count can be had by doing the insertion sort over the entire array
at the end, rather than across each small array as soon as the threshold is
reached. At the end of the partition sorting, LaMarca places a sentinel at the
end of the array and uses an insertion sort, eliminating the bounds check from
the end. In this implementation, the sentinel is placed at the front, and the
smallest of the first \cc{THRESHOLD} keys is used as a sentinel, in keeping
with our policy of not exceeding the bounds of the array.

@@

The number of keys in the linked list is 1022 since a word is needed for the
count, and another as a pointer to the next list. 1024 is the number of 32-bit
words which fit in a 4KB page of memory. In earlier versions where each page was
\cc{malloc}ed individually, the number of keys in the linked list was 1018,
as 4 words were used by the particular version of \cc{malloc} used by
SimpleScalar\footnote{This particular \cc{malloc} forms part of
\n{glibc}, the GNU C library, version 1.09.}.

### 2005-08-14 docs/tech_report/radix.tex
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage

All other sorts considered in this project are \n{comparison based} sorts.
They compare two keys at a time, and determine the larger and smaller key,
sorting on that basis. Radixsort is a \n{counting based} sort. It works by
separating the keys into small parts, and processing these parts without
regard for other keys. For radix of two, these small parts are bits, and for a
radix of 10, these parts would be digits.

@@

\subsection{Expected Performance}
Radixsort is expected to have very good performance, compared to comparison
based sorts. This is due to that fact that radixsort is $O(N)$; for an array as
small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will take 22 times
as many steps to complete. However, it is not expected that radixsort will
perform many times better than quicksort, the reigning champion of $O(NlogN)$
sorts. A `step' for quicksort is as simple as a comparison each step and a
possible exchange every few steps.  For radixsort, a step comprises being
counted four times and copied four times, with the counting involving several
layers of indirection, and causing a complete iteration across the array at the
start. LaMarca's results show the instruction count of quicksort and radixsort
being in the same order of magnitude, and we expect the same.

@@

The number of branch prediction misses should be exceptionally low. In
comparison based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. Radixsort has no comparative branches at all. As a result, only
flow control branches occur in radixsort, and they are all predictable.

>>>>

All other sorts considered in this project are \n{comparison-based} sorts.
They compare two keys at a time, and determine the larger and smaller key,
sorting on that basis. Radixsort is a \n{counting-based} sort. It works by
separating the keys into small parts, and processing these parts without
regard for other keys. For radix of two, these small parts are bits, and for a
radix of 10, these parts would be digits.

@@

\subsection{Expected Performance}
Radixsort is expected to have very good performance, compared to
comparison-based sorts. This is due to that fact that radixsort is $O(N)$; for
an array as small as 4096 records, it involves 12 times fewer steps than the
$O(NlogN)$ sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will
take 22 times as many steps to complete. However, it is not expected that
radixsort will perform many times better than quicksort, the reigning champion
of $O(NlogN)$ sorts. A `step' for quicksort is as simple as a comparison each
step and a possible exchange every few steps.  For radixsort, a step comprises
being counted four times and copied four times, with the counting involving
several layers of indirection, and causing a complete iteration across the array
at the start. LaMarca's results show the instruction count of quicksort and
radixsort being in the same order of magnitude, and we expect the same.

@@

The number of branch prediction misses should be exceptionally low. In
comparison-based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. Radixsort has no comparative branches at all. As a result, only
flow control branches occur in radixsort, and they are all predictable.
